Closing the empirical cycle

### 2022-11-03 action_letter.Rmd
update 2022-11-03 12:39:10

> Identify the predictor $k$ and value of that predictor that, when using that predictor and value to split the sample in two, maximizes the difference in model parameters between the two post-split groups (as determined by a likelihood ratio test).

@@

> Build a SEM-tree by repeating steps 3 and 4 until all post-split groups reach a minimum size that is too small for further splitting. The tree is then fully grown.

@@

> The SEM-forests algorithm is based on the random forests algorithm, which in turn is based on regression trees [see @stroblIntroductionRecursivePartitioning2009].
A single regression tree splits the sample repeatedly in two by whichever predictor $p$ makes both post-split groups maximally internally homogenous, and maximally different from one another.
A regression forest consists of many regression trees grown on bootstrap samples of the data. To ensure that each tree learns something different from the data, it may only consider a random subsample of $K$ candidates out of all $p$ predictors at each split.
With this in mind, the SEM-forest algorithm operates as follows... [etc]

@@

> Given an individual's scores on the predictor variables $p$, we can now traverse each tree in the forest to obtain a predicted value of the model parameters (in this case, the latent growth parameters). Aggregating the predicted parameters across all SEM-trees in the forest averages out prediction error.

@@

> As a simple example, imagine that only very low levels of autonomy support would be associated with elevated levels of emotion regulation (i.e., a higher intercept).
In regression analysis, this could be represented with a linear and quadratic term, and a plot might show that the effect of autonomy support on the intercept increases to a value $x_{knee}$, and then tapers off.
A tree-based model would represent the same effect by splitting the sample on autonomy support at the value $x_{knee}$.

@@

> Other limitations related to SEM-forests include the fact that,
unlike in traditional random forests, there is no error metric that could be used to calculate fit in a hold-out testing sample.
By extension, at the time of writing, there are not yet methods for cross-validation, model tuning, or evaluating the overall fit of a SEM-forest, as these too require an error metric.

@@

> A Reviewer suggested following up the planned machine learning analysis with follow-up analyses that might be more familiar to a developmental readership:
Regressing the parameters (intercept, slope and quadratic slope) of a latent growth model on each predictor in turn.
Such analyses might be helpful to plan hypotheses and perform power analysis for follow-up research within the SEM framework.
However, we strongly caution against over-interpretation.
Unlike the machine learning analyses, these analyses do not accommodate non-linear effects or interactions, nor are effects controlled for other predictors.
The results are merely descriptive and should not be used to perform inference.
[Supplemental Table S3](https://github.com/cjvanlissa/veni_forest/blob/master/supplemental_table_3.csv) reports standardized regression coefficients with significance asterisks for all continuous and binary predictors, and significance asterisks of Likelihood Ratio tests for multi-group categorical predictors.
All models had good fit ($RMSEA < .07$, $CFI > .99$, $SRMR < .04$).
Out of all predictors, 79% had a significant effect on the intercept, 24% on the slope, and 18% on the quadratic slope.
This indicates that our machine learning analyses indeed recovered relevant patterns in the data.

@@

> The source code is available at <https://github.com/cjvanlissa/shiny_errata>;
users can run the app from a local R instance,
or view an online demo using a link in the README file of this repository (as long hosting is available).

@@

@@

We apologize. The manuscript only has 42 pages; pages 43-65 are copies of the online supplemental materials. The submission portal has trouble correctly presenting these materials, but they will not be part of the paper.

@@

@@

@@

@@

@@

We understand that pages 43-65 are not very useful in their current form, and apologize for the inconvenience. We were required to attach hard copies of the online supplemental tables during submission, and these were evidently not correctly formatted by the submission portal. They are just Excel spreadsheets, and will not be part of the manuscript.

>>>>

`r get_revision(man, "stepiii")`

@@

`r get_revision(man, "fullygrown")`

@@

@@

`r get_revision(man, "predictions")`

@@

`r get_revision(man, "examplenonlinear")`

@@

`r get_revision(man, "limitations")`

@@

`r get_revision(man, "posthoc")`

@@

`r get_revision(man, "errata")`

@@

We thank the Reviewer for this hepful suggestion, and have written the following to address it:

`r get_revision(man, "includepersonality")`

@@

We apologize. The manuscript only has 42 pages; pages 43-65 are copies of the online supplemental materials (which are supposed to be accessed via the broken links, which have now been fixed). For the sake of blind review, we also submitted copies of these tables through the submission portal, but it evidently cannot correctly present these materials. The tables will not be part of the paper in this form.

@@

We understand why the Reviewer expects the analysis to include model tuning and cross-validation; these are standard procedures in machine learning.
However, these methods do not (yet?) exist for SEM-forests.
This is because there is no single error metric for SEM-forests that can be optimized.
For standard random forests, this error metric is the root mean squared error (difference between predicted and observed values for all cases).
The "outcome" of each regression tree in the SEM-forest is an entire SEM-model, for which there is no clear-cut case-specific error metric.
It is important to note, however, that unlike other techniques, random forests typically achieve competitive performance even with default settings [@probstHyperparametersTuningStrategies2019].
Thus, even when a model isn't tuned - or cannot be tuned, as is the case here - it should perform well.
Although we did attempt to explain these points in the previous version of the manuscript,
we have attempted to address this comment by further emphasizing this in the Revision:

`r get_revision(man, "limitations")`

@@

We thank the Reviewer for this excellent suggestion, and now provide this figure

@@

We appreciate this point, and the writing here should certainly be more nuanced (that was an oversight).
We have rephrased this as follows:

`r get_revision(man, "riskassessment")`

@@

As requested, we now provide more information about cultural values that may impact the generalizability of our findings:

`r get_revision(man, "culture")`

@@

We apologize for the inconvenience. Pages 43-65 are not intended to be used; we were required to attach hard copies of the online supplemental tables in the submission portal, and these were evidently not formatted correctly. They are just spreadsheets, available online, and will not be part of the printed manuscript.

### 2022-11-03 manuscript/manuscript.Rmd
update 2022-11-03 12:39:10

1. Bootstrap the data with replacement; we used 1000 bootstrap replications
2. For each bootstrap sample, construct a SEM tree:
    i. Estimate the latent growth model on the bootstrap sample
    ii. Consider a random selection of $K$ candidates out of all $p$ predictors; we used $k = 9$ based on the standard practice to use $K = \lfloor\sqrt{p}\rfloor$
    iii. Identify the predictor $k$ and value of that predictor that, when using that predictor and value to split the sample in two, maximizes the difference in model parameters between the two post-split groups (as determined by a likelihood ratio test).
    iv. Build a SEM-tree by repeating steps 3 and 4 until all post-split groups reach a minimum size that is too small for further splitting. The tree is then fully grown.
3. Given an individual's scores on the predictor variables $p$, we can now traverse each tree in the forest to obtain a predicted value of the model parameters (in this case, the latent growth parameters). Aggregating the predicted parameters across all SEM-trees in the forest averages out prediction error.

Each tree thus maps relationships between the predictors and outcome as a series of recursive splits of the data.
This approach intrinsically accommodates non-linear effects and (complex) interactions, which is an advantage over parametric models that assume linearity.
As a simple example, imagine that only very low levels of autonomy support would be associated with elevated levels of emotion regulation (i.e., a higher intercept).
In regression analysis, this could be represented with a linear and quadratic term, and a plot might show that the effect of autonomy support on the intercept increases to a value $x_{knee}$, and then tapers off.
A tree-based model would represent the same effect by splitting the sample on autonomy support at the value $x_{knee}$.
The algorithm additionally performs variable selection;
unimportant variables are not selected in step 2.iii.
Thus, the number of predictors can be large without risking model non-identification.</span>

@@

A Reviewer suggested following up the planned machine learning analysis with follow-up analyses that might be more familiar to a developmental readership:
Regressing the parameters (intercept, slope and quadratic slope) of a latent growth model on each predictor in turn.
Such analyses might be helpful to plan hypotheses and perform power analysis for follow-up research within the SEM framework.
However, we strongly caution against over-interpretation.
Unlike the machine learning analyses, these analyses do not accommodate non-linear effects or interactions, nor are effects controlled for other predictors.
The results are merely descriptive and should not be used to perform inference.
[Supplemental Table S3](https://github.com/cjvanlissa/veni_forest/blob/master/supplemental_table_3.csv) reports standardized regression coefficients with significance asterisks for all continuous and binary predictors, and significance asterisks of Likelihood Ratio tests for multi-group categorical predictors.
All models had good fit ($RMSEA < .07$, $CFI > .99$, $SRMR < .04$).
Out of all predictors, 79% had a significant effect on the intercept, 24% on the slope, and 18% on the quadratic slope.
This indicates that our machine learning analyses indeed recovered relevant patterns in the data.

@@

@@

These findings have several implications for the early identification of youth at-risk for emotion regulation difficulties.
The fact that individual-level predictors were most important
suggests that individual assessments might be sufficient.
Holistic approaches that involve the family system might not be necessary for preliminary risk assessment.
Moreover, some of the most important predictors are already routinely assessed (for example, big five personality).
Other important predictors are overt, and thus easily observed by teachers and parents - for example, externalizing problem behavior and conflict frequency and behavior.
Future research might thus develop and validate risk assessment instruments
focusing on these constructs.

The method used in the present study is particularly well-suited to risk assessment instruments,
because machine learning models maximize predictive performance, whereas traditional methods like linear regression maximize performance within the study sample.
In fact, the trained model from the present study can be used to make predictions for the emotion regulation development of new adolescents.
To illustrate the potential for risk assessment, we developed ERRATA,
an Emotion Regulation Risk Assessment Tool for Adolescents (see Figure \@ref(fig:figerrata)).
Users can select values of categorical predictors,
and dial in low- to high values on a five point scale for each continuous predictor,
which are mapped to the percentiles of the observed data distribution corresponding to -2SD, -1SD, the median, +1SD, and +2SD of a standard normal distribution.
By default, all inputs are set to the median and mode values.
This means that users can use informative input for the most important predictors,
while keeping other less important predictors constant at the most common value.
The ERRATA application is open source, and can thus be customized for clinical use.
The source code is available at <https://github.com/cjvanlissa/shiny_errata>;
users can run the app from a local R instance,
or view an online demo using a link in the README file of this repository (as long hosting is available).

@@

The study also has limitations, most notably that the sample was selective,
which may limit the generalizability of the findings.
Specifically, this was a non-random, ethnically and socially relatively homogeneous sample from the Netherlands.
This raises specific concerns, for example, that the low importance of the demograpic factors SES and religion might be due to a restriction of range.
The sample might also be somewhat dated - especially considering daily life has changed drastically for youth coming of age in a post-pandemic world.
The only way to address these concerns is to replicate these findings in other samples.
Note that all code is made available in a reproducible archive to facilitate replication.
Another limitation is that the data source was a panel study with mostly questionnaire measures.
All conclusions are conditional on the available data; thus, for example,
our study cannot speak to the potential importance of biological (except pubertal development),
cognitive and neurological factors.
It is also important to consider that, in our efforts to identify important early predictors of emotion dysregulation, we relied on predictors at a single timepoint (age 13).
This does not account for over time change in the relative importance of predictors.
Another potential limitation is common method bias.
Many predictors were assessed using adolescent-reported questionnaire measures,
which raises the question whether individual-level predictors ranked highly due to shared method variance.
This concern is assuaged somewhat by the observation that parent-reported externalizing behavior (5) was much more predictive than adolescent self-reported externalizing behavior (30).
This indicates that predictors less susceptible to common method bias nonetheless explained substantial variance.
One final limitation related to the data is the relatively small sample size.
Machine learning methods benefit from large samples.
Although random forests are relatively robust to overfitting,
larger samples will likely yield more generalizable results.
<span i = "homogeneity2">Larger samples also afford more complicated models.
We saw the need, for example, to assume homogeneity of residuals within each node of the SEM-forest.
This assumption is likely violated in some nodes; a larger sample would allow us to freely estimate residuals.
Note, however, that many other assumptions are also likely violated in some nodes (e.g., the shape of the growth trajectory).
This limitation is inherent to the method.</span>
Other limitations related to SEM-forests include the fact that,
unlike in traditional random forests, there is no error metric that could be used to calculate fit in a hold-out testing sample.
By extension, at the time of writing, there are not yet methods for cross-validation, model tuning, or evaluating the overall fit of a SEM-forest, as these too require an error metric.
Finally, SEM-forests require substantial computational resources,
which also limit the ability to compute different iterations of the model beyond default settings,
or to exploring marginal effects for higher order interactions.

>>>>

1. Bootstrap the data with replacement; we used 1000 bootstrap replications
2. For each bootstrap sample, construct a SEM tree:
    i. Estimate the latent growth model on the bootstrap sample
    ii. Consider a random selection of $K$ candidates out of all $p$ predictors; we used $k = 9$ based on the standard practice to use $K = \lfloor\sqrt{p}\rfloor$
    iii. [Identify the predictor $k$ and value of that predictor that, when using that predictor and value to split the sample in two, maximizes the difference in model parameters between the two post-split groups (as determined by a likelihood ratio test).]{#stepiii}
    iv. [Build a SEM-tree by repeating steps 3 and 4 until all post-split groups reach a minimum size that is too small for further splitting. The tree is then fully grown.]{#fullygrown}
3. [Given an individual's scores on the predictor variables $p$, we can now traverse each tree in the forest to obtain a predicted value of the model parameters (in this case, the latent growth parameters). Aggregating the predicted parameters across all SEM-trees in the forest averages out prediction error.]{#predictions}

Each tree thus maps relationships between the predictors and outcome as a series of recursive splits of the data.
This approach intrinsically accommodates non-linear effects and (complex) interactions, which is an advantage over parametric models that assume linearity.
<span id = "examplenonlinear">As a simple example, imagine that only very low levels of autonomy support would be associated with elevated levels of emotion regulation (i.e., a higher intercept).
In regression analysis, this could be represented with a linear and quadratic term, and a plot might show that the effect of autonomy support on the intercept increases to a value $x_{knee}$, and then tapers off.
A tree-based model would represent the same effect by splitting the sample on autonomy support at the value $x_{knee}$.</span>
The algorithm additionally performs variable selection;
unimportant variables are not selected in step 2.iii.
Thus, the number of predictors can be large without risking model non-identification.</span>

@@

<span id = "posthoc">A Reviewer suggested following up the planned machine learning analysis with follow-up analyses that might be more familiar to a developmental readership:
Regressing the parameters (intercept, slope and quadratic slope) of a latent growth model on each predictor in turn.
Such analyses might be helpful to plan hypotheses and perform power analysis for follow-up research within the SEM framework.
However, we strongly caution against over-interpretation.
Unlike the machine learning analyses, these analyses do not accommodate non-linear effects or interactions, nor are effects controlled for other predictors.
The results are merely descriptive and should not be used to perform inference.
[Supplemental Table S3](https://github.com/cjvanlissa/veni_forest/blob/master/supplemental_table_3.csv) reports standardized regression coefficients with significance asterisks for all continuous and binary predictors, and significance asterisks of Likelihood Ratio tests for multi-group categorical predictors.
All models had good fit ($RMSEA < .07$, $CFI > .99$, $SRMR < .04$).
Out of all predictors, 79% had a significant effect on the intercept, 24% on the slope, and 18% on the quadratic slope.
This indicates that our machine learning analyses indeed recovered relevant patterns in the data.</span>

@@

<span id = "includepersonality">Several personality traits emerged as particularly important proximal predictors.
According to a recent review on personality and emotion regulation (in adults),
personality is a broad framework for describing psychological capacities,
of which emotion regulation is one [@hughesPersonalityTraitsEmotion2020].
This view is certainly supported by our findings,
and yet research on adolescents does not routinely assess the role of personality alongside factors of particular interest in our discipline, such as parenting, peer relationships, and problem behavior [@vanlissaMappingPhenomenaRelevant2021].
This is certainly a promising area for future research.</span>

@@

[These findings have several implications for the early identification of youth at-risk for emotion regulation difficulties.
The fact that individual-level predictors were most important in predicting adolescents' diverging developmental trajectories suggests that it might be possible to perform preliminary risk assessments for emotion dysregulation using only adolescent-reported individual differences.
Although the focus in clinical practice is shifting towards holistic approaches that involve the family system,
this may not always be feasible.
For instance, a family member may refuse to cooperate, or holistic assessment may be too costly.
In such cases, it is useful to know that individual differences carry substantial informational value.]{#riskassessment}
Moreover, some of the most important predictors are already routinely assessed (for example, big five personality).
Other important predictors are overt, and thus easily observed by teachers and parents - for example, externalizing problem behavior and conflict frequency and behavior.
Future research might thus develop and validate risk assessment instruments
focusing on these constructs.

The method used in the present study is particularly well-suited to risk assessment instruments,
because machine learning models maximize predictive performance, whereas traditional methods like linear regression maximize performance within the study sample.
In fact, the trained model from the present study can be used to make predictions for the emotion regulation development of new adolescents.
To illustrate the potential for risk assessment, we developed ERRATA,
an Emotion Regulation Risk Assessment Tool for Adolescents (see Figure \@ref(fig:figerrata)).
Users can select values of categorical predictors,
and dial in low- to high values on a five point scale for each continuous predictor,
which are mapped to the percentiles of the observed data distribution corresponding to -2SD, -1SD, the median, +1SD, and +2SD of a standard normal distribution.
By default, all inputs are set to the median and mode values.
This means that users can use informative input for the most important predictors,
while keeping other less important predictors constant at the most common value.
<span id = "errata">The ERRATA application is open source, and can thus be customized for clinical use.
The source code is available at <https://github.com/cjvanlissa/shiny_errata>;
users can run the app from a local R instance,
or view an online demo using a link in the README file of this repository (as long hosting is available).</span>

@@

The study also has limitations, most notably that the sample was selective,
which may limit the generalizability of the findings.
Specifically, this was a non-random, ethnically and socially relatively homogeneous sample from the Netherlands.
[Cultural values, which differ across and within countries,
have been found to influence people's motivation to regulate emotions, as well as the strategies they use and their adaptiveness [@fordCultureEmotionRegulation2015].
In cross-cultural comparisons, the Netherlands has been found (on average) to ascribe to similar cultural values as large Anglophonic countries (e.g., the United States, Canada, and Australia) [@terlutterGLOBEStudyApplicability2006].
Whether our findings also generalize to populations that adhere to different cultural values remains an open question.]{#culture}
Another concern related to generalizability is that the relative homogeneity of our sample in terms of demograpic factors, like SES and religion, might have contributed to their lower variable importance.
The sample might also be somewhat dated - especially considering daily life has changed drastically for youth coming of age in a post-pandemic world.
The only way to address these concerns is to replicate these findings in other samples.
To facilitate such replications, all code is made available in a reproducible archive.
Another limitation is that the data source was a panel study with mostly questionnaire measures.
All conclusions are conditional on the available data; thus, for example,
our study cannot speak to the potential importance of biological (except pubertal development),
cognitive and neurological factors.
It is also important to consider that, in our efforts to identify important early predictors of emotion dysregulation, we relied on predictors at a single timepoint (age 13).
This does not account for over time change in the relative importance of predictors.
Another potential limitation is common method bias.
Many predictors were assessed using adolescent-reported questionnaire measures,
which raises the question whether individual-level predictors ranked highly due to shared method variance.
This concern is assuaged somewhat by the observation that parent-reported externalizing behavior (5) was much more predictive than adolescent self-reported externalizing behavior (30).
This indicates that predictors less susceptible to common method bias nonetheless explained substantial variance.
One final limitation related to the data is the relatively small sample size.
Machine learning methods benefit from large samples.
Although random forests are relatively robust to overfitting,
larger samples will likely yield more generalizable results.
<span id = "homogeneity2">Larger samples also afford more complicated models.
We saw the need, for example, to assume homogeneity of residuals within each node of the SEM-forest.
This assumption is likely violated in some nodes; a larger sample would allow us to freely estimate residuals.
Note, however, that many other assumptions are also likely violated in some nodes (e.g., the shape of the growth trajectory).
This limitation is inherent to the method.</span>
<span id = "limitations">Other limitations related to SEM-forests include the fact that,
unlike in traditional random forests, there is no error metric that could be used to calculate fit in a hold-out testing sample.
By extension, at the time of writing, there are not yet methods for cross-validation, model tuning, or evaluating the overall fit of a SEM-forest, as these too require an error metric.
Note, however, that even without model tuning, random forests tend to achieve competitive performance [@probstHyperparametersTuningStrategies2019].</span>
Finally, SEM-forests require substantial computational resources,
which also limit the ability to compute different iterations of the model beyond default settings,
or to exploring marginal effects for higher order interactions.
